### 2017-12-12 submission/manuscript.Rmd
add results section models --better version

***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
We then tested whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone. To test this we used two approaches. The first used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera. The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model across studies using either all OTUs or only OTUs that taxonomically classified to CRC-associated genera.

The genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3]. With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6]. The AUC for classification of adenomas was lower than carcionmas for both tissue and stool [Figure S4-S6]. Interestingly, the AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].

The OTU based models for both fecal and tissue (matched and unmatched) samples, showed an AUC decrease when only OTUs from the CRC-associated genera are used versus the full community of OTUs [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5] with the largest difference in median AUC for stool being in carcinoma classification [Figure 4B] and for tissue being an adenoma [Figure 5A].

In stool the most common genera in the genus based models belonged mostly to resident genera [Figure 6A & B]. This included genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia*. With respect to the CRC-associated genera *Fusobacterium* was the only genus present in adenoma while all four were present in carcinoma [Figure 6A & B]. However, none of these CRC-associated genera were present in the majority of studies. For the adenoma OTU models, OTUs that classified as  *Ruminococcaceae* or *Roseburia* were present in the top 10 OTUs for the vast majority of studies [Figure 6C]. Although for the carcinoma OTU models *Ruminococcaceae* was in the top 10 for many studies, *Bacteroides* was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].

Conversely to the stool models, both genera and OTU based models in tissue had the vast majority of their top 10 occur in a study specific manner [Figure S7]. *Fusobacterium* and *Fusobacteriaceae* show up more often in the top 10 for matched tissue samples but are not present in the top 10 or much lower for unmatched tissue [Figure S7B-C & S7E-F]. There appears to be very little overlap in the top 10 most important variables between stool and tissue for both adenoma and carcinoma [Figure 6 & S7].

>>>>

***Whole Community Models Add Important Community Context:***
Since CRC-associated genera had increased RR for carcinoma over diversity metrics we wondered whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone. To test this we used two approaches. The first approach used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera. The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model across studies using either all OTUs or only OTUs that taxonomically classified to CRC-associated genera.

Our first approach using genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3]. With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6]. Two interesting but seperate general observations from these models were that: 1) classification of adenomas was lower than carcinomas for both tissue and stool and 2) AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].

A similar trend was observed for the OTU based models for both fecal and tissue (matched and unmatched) samples. There was a generalized decrease in AUC when only OTUs from the CRC-associated genera were used versus the full community of OTUs for both adenoma and carcinoma [Figure 4 & 5]. The largest difference in median AUC for stool, between the full community of OTUs and only CRC-associated OTUs, was in carcinoma classification [Figure 4B] and for tissue it was in adenoma classification [Figure 5A].

In stool the most common genera in the genus based models belonged mostly to resident genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. With respect to the CRC-associated genera, *Fusobacterium* was the only genus present in adenoma while all four were present in carcinoma [Figure 6A & B]. Conversely, none of these CRC-associated genera were present in the majority of studies. When we move onto the OTU based models, the adenoma OTU models had OTUs that classified as  *Ruminococcaceae* or *Roseburia* present in the top 10 OTUs for the vast majority of studies [Figure 6C]. *Ruminococcaceae* was also present in the top 10 in some studies for the carcinoma OTU models, but it was *Bacteroides* that was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].

Conversely to the stool models, both genera and OTU based models in tissue had the vast majority of their top 10 occur in a study specific manner [Figure S7]. *Fusobacterium* and *Fusobacteriaceae* show up more often in the top 10 for matched tissue samples but were not present in the top 10 or were much lower ranked for unmatched tissue [Figure S7B-C & S7E-F]. There appears to be very little overlap in the top 10 most important variables between stool and tissue for both adenoma and carcinoma [Figure 6 & S7].

### 2017-12-12 submission/manuscript.tex
add results section models --better version

\textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
Genera Increases Model AUC:}} We then tested whether the overall
bacterial community was at all important to classifying disease or if
the CRC-associated genera were sufficient alone. To test this we used
two approaches. The first used genus level data and tested whether there
were any differences in AUC when training on one study and testing on
all the others when using either all genera present or only the
CRC-associated genera. The second approach used OTU level data and
tested whether there was a generalized decrease in the 10-fold cross
validation (CV) model across studies using either all OTUs or only OTUs
that taxonomically classified to CRC-associated genera.

The genus based models showed an AUC decrease in model classification on
the training set for both stool and tissue studies {[}Figure S2-S3{]}.
With respect to the test sets, comprised of genera data from other
studies, both the all genera model and CRC-associated models had a
similar ability to detect adenomas or carcinomas {[}Figure S4-S6{]}. The
AUC for classification of adenomas was lower than carcionmas for both
tissue and stool {[}Figure S4-S6{]}. Interestingly, the AUC for the
classification of carcinoma was consistently lower for the tissue models
than the stool models {[}Figure S4-S6{]}.

The OTU based models for both fecal and tissue (matched and unmatched)
samples, showed an AUC decrease when only OTUs from the CRC-associated
genera are used versus the full community of OTUs {[}Figure 4 \& 5{]}.
This decrease is observed in both adenoma and carcinoma groups {[}Figure
4 \& 5{]} with the largest difference in median AUC for stool being in
carcinoma classification {[}Figure 4B{]} and for tissue being an adenoma
{[}Figure 5A{]}.

In stool the most common genera in the genus based models belonged
mostly to resident genera {[}Figure 6A \& B{]}. This included genera
such as \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}.
With respect to the CRC-associated genera \emph{Fusobacterium} was the
only genus present in adenoma while all four were present in carcinoma
{[}Figure 6A \& B{]}. However, none of these CRC-associated genera were
present in the majority of studies. For the adenoma OTU models, OTUs
that classified as \emph{Ruminococcaceae} or \emph{Roseburia} were
present in the top 10 OTUs for the vast majority of studies {[}Figure
6C{]}. Although for the carcinoma OTU models \emph{Ruminococcaceae} was
in the top 10 for many studies, \emph{Bacteroides} was present in the
overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.

Conversely to the stool models, both genera and OTU based models in
tissue had the vast majority of their top 10 occur in a study specific
manner {[}Figure S7{]}. \emph{Fusobacterium} and \emph{Fusobacteriaceae}
show up more often in the top 10 for matched tissue samples but are not
present in the top 10 or much lower for unmatched tissue {[}Figure S7B-C
\& S7E-F{]}. There appears to be very little overlap in the top 10 most
important variables between stool and tissue for both adenoma and
carcinoma {[}Figure 6 \& S7{]}.

>>>>

\textbf{\emph{Whole Community Models Add Important Community Context:}}
Since CRC-associated genera had increased RR for carcinoma over
diversity metrics we wondered whether the overall bacterial community
was at all important to classifying disease or if the CRC-associated
genera were sufficient alone. To test this we used two approaches. The
first approach used genus level data and tested whether there were any
differences in AUC when training on one study and testing on all the
others when using either all genera present or only the CRC-associated
genera. The second approach used OTU level data and tested whether there
was a generalized decrease in the 10-fold cross validation (CV) model
across studies using either all OTUs or only OTUs that taxonomically
classified to CRC-associated genera.

Our first approach using genus based models showed an AUC decrease in
model classification on the training set for both stool and tissue
studies {[}Figure S2-S3{]}. With respect to the test sets, comprised of
genera data from other studies, both the all genera model and
CRC-associated models had a similar ability to detect adenomas or
carcinomas {[}Figure S4-S6{]}. Two interesting but seperate general
observations from these models were that: 1) classification of adenomas
was lower than carcinomas for both tissue and stool and 2) AUC for the
classification of carcinoma was consistently lower for the tissue models
than the stool models {[}Figure S4-S6{]}.

A similar trend was observed for the OTU based models for both fecal and
tissue (matched and unmatched) samples. There was a generalized decrease
in AUC when only OTUs from the CRC-associated genera were used versus
the full community of OTUs for both adenoma and carcinoma {[}Figure 4 \&
5{]}. The largest difference in median AUC for stool, between the full
community of OTUs and only CRC-associated OTUs, was in carcinoma
classification {[}Figure 4B{]} and for tissue it was in adenoma
classification {[}Figure 5A{]}.

In stool the most common genera in the genus based models belonged
mostly to resident genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} {[}Figure 6A \& B{]}. With
respect to the CRC-associated genera, \emph{Fusobacterium} was the only
genus present in adenoma while all four were present in carcinoma
{[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
were present in the majority of studies. When we move onto the OTU based
models, the adenoma OTU models had OTUs that classified as
\emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
for the vast majority of studies {[}Figure 6C{]}. \emph{Ruminococcaceae}
was also present in the top 10 in some studies for the carcinoma OTU
models, but it was \emph{Bacteroides} that was present in the
overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.

Conversely to the stool models, both genera and OTU based models in
tissue had the vast majority of their top 10 occur in a study specific
manner {[}Figure S7{]}. \emph{Fusobacterium} and \emph{Fusobacteriaceae}
show up more often in the top 10 for matched tissue samples but were not
present in the top 10 or were much lower ranked for unmatched tissue
{[}Figure S7B-C \& S7E-F{]}. There appears to be very little overlap in
the top 10 most important variables between stool and tissue for both
adenoma and carcinoma {[}Figure 6 \& S7{]}.
